Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC). }

### 2015-06-05 tex/body.tex
more

Our setting is a monomorphic (or many-sorted) first-order logic
like the one defined by SMT-LIB \cite{smtlib25}.
A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of %ct type symbols
first-order types
%
%\ct{What the rationale for using the letter $y$ in $\stypes{\Sigma}$?}
%\jb{y as in t\textbf{y}pe. changed to `ty'}
(or sorts) and a set $\sfuns{\Sigma}$ of function symbols over these types.
We assume that signatures always contain a Boolean type $\Bool$ and constants
$\ltrue, \lfalse : \Bool$ for truth and falsity,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard Boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and an if--then--else function symbol
$\lite : \Bool \times \tau \times \tau \rightarrow \tau$
for each $\tau \in \stypes{\Sigma}$.
%
For each $\tau \in \stypes{\Sigma}$,
we fix an infinite set $\svars{\Sigma}_\tau$ of \emph{variables of type $\tau$} and
define $\svars{\Sigma}$ as $\bigcup_{\tau \in \stypes{\Sigma}} \svars{\Sigma}_\tau$.
$\Sigma$-terms are built as usual over functions symbols in $\Sigma$ and variables in $\svars{\Sigma}$.
%
Formulas are terms of type $\Bool$.
We write $t\typ{\tau}$ to denote a metavariable ranging over terms of
type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
Given a term $u$, we write $u[\vec t/\vec x]$ to denote the result of replacing
all occurrences of $\vec x$ with $\vec t$ in $u$.
\begin{longv}When applied to terms, the symbol $=$ denotes syntactic equality.\end{longv}

@@

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well-founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)$
where $T$ is integer arithmetic extended with the uninterpreted symbol 
$\con p:\typeint \rightarrow \Bool$, can be shown well-founded under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. \begin{longv}Logically, such dependencies can be captured by
congruence rules. \end{longv}Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

@@

%\newcommand\badassex{
% \{\forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x ),\;\,
% \forallf{\con{g}} x : \typeint.\;\, \lite( \con{f}( x ) \teq 5,\allowbreak\;
%   \con{g}( x ) \teq \con{f}( x ),\allowbreak\;
%   \con{g}( x ) \teq \con{g}( x ) + 1)\}}
\newcommand\badassex{
 \{\forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x ),\;\,
 \forallf{\con{g}} x : \typeint.\;\, \con{g}( x ) \teq \con{g}( x ) + \nobreak\con{f}( x )\}}

An example of an inadmissible set is
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$,
where $T$ is integer arithmetic extended to a set of uninterpreted symbols 
$\{\con f:\typeint \rightarrow \typeint,\, \con g:\typeint \rightarrow \typeint,\, \ldots \}$.
The reason is that the formula $\ltrue$ is (trivially) closed under function expansion with respect to this set,
and there is no model of $T$
satisfying $\con{f}$'s definition. A more subtle example is
\begin{shortv}$\badassex$. \end{shortv}%
\begin{longv}\[\badassex\]\end{longv}%
While this set has a model where $\con f$ and $\con g$ are interpreted as the
constant function $0$, it is not admissible since %the formula
$\con{f}( 0 ) \teq 1$ is
closed under function expansion \begin{longv}with respect to this set\end{longv}
and yet there exists
no interpretation satisfying both $\con{f}( 0 ) \teq 1$ and $\con{g}$'s
definition.

@@

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
%\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{if}$ $\tau = \Bool$ $\mathsf{and}$ $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$%\kern6.35em$\mid$
      \itemx[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \itemx[] $\mathsf{if}$ $p = \ppos$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ $\mathsf{if}$ $p = \pneg$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
%\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
%\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
%\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\mathsf{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec t \mid \con{f}( \vec t ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
%\\[-.8\baselineskip]
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$ $\mathsf{let}$ $( \varphi',\, \_ ) = \conv_0( \varphi,\, \ppos )$ $\mathsf{in}$ $\varphi'$
%\\[-1.25\baselineskip] %% TYPESETTING
\end{enumerate}
\caption{\,Definition of translation $\conv$}
\label{fig:A}
\end{figure}

We fix a signature $\Sigma$ and a $\Sigma$-formula
$\varphi$ in definitional form with respect to $\sfundefs{\Sigma}$,
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\phi'$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
We present a method that constructs an extended signature
$\extendsig{ \Sigma }$ and an $\extendsig{ \Sigma }$-formula $\varphi'$ such that
$\varphi'$ is satisfiable if and only if $\varphi$ is satisfiable---i.e.,
$\varphi$ and $\varphi'$ are \emph{equisatisfiable}.
The idea behind this translation
is to use an uninterpreted type to abstract the set of
\relax{relevant} tuples for each defined function $\con{f}$ and restrict the
quantification of $\con{f}$'s definition to a variable of this
type. Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.

More precisely,
suppose our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
For each $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau \in \sfundefs{\Sigma}$,
the extended signature $\extendsig{\Sigma}$ contains
%\begin{itemize}
%\item
an uninterpreted \emph{abstract type} $\fargtype{\con{f}}$ and
%\item
$n$ uninterpreted \emph{concretization functions} $\fargx{\con{f}}{1} : \fargtype{\con{f}} \rightarrow \tau_1$, \ldots, $\fargx{\con{f}}{n} : \fargtype{\con{f}} \rightarrow \tau_n$.
%\end{itemize}
%The interpretation of uninterpreted type $\fargtype{\con{f}}$ will denote the elements (tuples) on which the function $\con{f}$ is applied.
%The role of the uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ will be discussed more in the following.

The translation $\conv$ defined in Figure~\ref{fig:A} translates the $\Sigma$-formula
$\varphi$ into an $\extendsig{\Sigma}$-formula. It relies
on the auxiliary function $\conv_0$, which takes two arguments:\ the term $t$
to translate and a \relax{polarity}~$p$, which is either $\ppos$, $\pneg$, or
$\pnone$. \vthinspace$\conv_0$ returns a pair $( t'\negvthinspace,\, \chi )$, where $t'$ is a term of
the same type as $t$ and $\chi$ is a $\extendsig{\Sigma}$-formula.
%an additional constraint.
%This constraint is needed if the polarity is $\pnone$; otherwise, it is encoded
%directly in $t'$ and $\chi$ is simply $\ltrue$.
%In the auxiliary function $\conv_0$, the function $\pol(
%\boolop, i, p )$ returns the polarity of the $i$\vvthinspace{th} argument of an
%application of~$\boolop$ having polarity $p$ if $\boolop$ is an interpreted
%predicate or $\pnone$ otherwise.

@@

The translation alters $\varphi$ in two main ways. First, it restricts the
quantification on function definitions for $\con{f}$ to the corresponding
uninterpreted type $\fargtype{\con{f}}$ and inserts appropriate selectors
$\fargx{\con{f}}{i}$. Second, it augments $\varphi$ with additional existential
constraints
$\exists a : \fargtype{\con{f}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec t$; these
ensure that the restricted definition for $\con{f}$ covers all relevant tuples
of terms, namely the ones for which an $\con{f}$ application exists and is
relevant to the satisfiability of $\varphi$.
%\jb{what does ``current branch of'' mean? can we drop it / rephrase it / vague
%it up?}

If $t$ is an application of a predicate $\boolop$, including the built-in
operators ${\oldneg}$, ${\wedge}$, ${\vee}$, ${\teq}$, and ${\lite}$,
$\conv_0$ calls itself recursively on the arguments $t_i$ and polarity $\pol(
\boolop, i, p )$, where $\pol$ is defined as
\[\pol( \boolop, i, p ) =
\begin{cases}
p & \text{if either $\boolop \in \{{\wedge}, {\vee}\}$ or $\boolop = \lite$ and $i \in \{2, 3\}$} \\[-\jot]
\pnegate p & \text{if $\boolop = {\neg}$} \\[-\jot]
\pnone & \text{otherwise}
\end{cases}\]
where the negation $\pnegate p$ of a polarity $p$ is $\pneg$ if $p$ is $\ppos$, $\ppos$
if $p$ is $\pneg$, and $\pnone$ if $p$ is $\pnone$.
If $t$ is positive, $\conv_0$ adds the constraint $\chi$ derived from
the subterms as a conjunction. Dually, if $t$ is negative, it constructs a
disjunction with the negated constraint, to achieve the same overall effect as
in the positive case. It $t$ has no polarity, the constraint $\chi$ is
returned to the caller.
%
\begin{longv}\par\end{longv}
%
If $t$ is a function definition, % of a function in $\sfundefs{\Sigma}$,
$\conv_0$ constructs a quantified formula over a single variable
$a$ of type $\fargtype{\con{f}}$ and replaces all occurrences of $\vec x$ in
$t'_1$ with $\vecfarg{\con{f}}( a )$. Since function definitions
are a top-level conjunct, %by case analysis on the return values of $\conv$,
$\chi$ must be $\ltrue$.
If $t$ is an unannotated quantified formula, $\conv_0$ calls itself
on the body with the same polarity. \begin{longv}A quantifier is prefixed to the
quantified formula and constraint returned by the recursive call.\end{longv}
Otherwise, $t$ is either an application of an uninterpreted predicate or a term
of a type other than $\Bool$. The returned constraint is a conjunction of
formulas of the form $\exists a : {\fargtype{\con{f}}}.\;\, \vecfarg{\con{f}}(
a ) \teq \vec t$ for each subterm $\con{f}( \vec t )$ of $t$ such that $\con{f}
\in \sfundefs{\Sigma}$. Such constraints, when asserted positively, ensure that
the tuple of uninterpreted functions $\vecfarg{\con{f}}$ has $\vec t$
in its range.

@@

\begin{example}
\label{ex:translation}
Let $T$ be linear arithmetic with the uninterpreted symbols $\{ \con{s} : {
\typeint \rightarrow \typeint },\, \const{c} : { \typeint } \}$.
Let $\varphi$ be the $\Sigma$-formula
\begin{equation} \label{eq:ex-before}
\forallf{\con{s}} x : {\typeint}.\;\, \lite\bigl( x \leq 0,\; \con{s}(x) \teq 0,\;
  \con{s}( x ) \teq x + \con{s}( x - 1 ) \bigr) \wedge \con{s}( \con{c} ) > 100
\end{equation}
%
The definition of $\const{s}$ specifies that it returns the sum of all
positive integers up to $x$. The formula $\varphi$ is in definitional form with
respect to $\sfundefs{\Sigma}$ 
%\jb{correct? It said $\Sigma_u$, which was definitely wrong} 
and states that the sum of all
positive numbers up to $\const{c}$ is greater than $100$. This formula is
satisfiable with a model that interprets $\const{c}$ as $14$ or more.
Due to the universal quantifier,
current SMT techniques
%~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13,reynolds-et-al-2013}
%%% We described them in the intro. When I see references like that appear with
%%% no description, as a reader, I get all paranoid and wonder if it's the old
%%% stuff from the intro (Ge, Moura, Renoylds et al.) or if it's new stuff ---
%%% and I cannot remember numbers like [9,18,19] across sections or even
%%% paragraphs. --jb
are unable to find
a model for this formula. The signature $\extendsig{\Sigma}$ contains the type
$\fargtype{\con{s}}$ and the uninterpreted function $\farg{\con{s}} : \fargtype{\con{s}}
\rightarrow \typeint$. The result of $\conv( \varphi )$ 
is $\varphi'$, which (after simplification) is the $\extendsig{\Sigma}$-formula
%
\begin{equation} \label{eq:ex-after}
\!\begin{aligned}[c]
  & \phantom{{\wedge}\; }\bigl(
      \forall a : \fargtype{\con{s}}.\; \lite\bigl(
        \!\begin{aligned}[t]
         &  \farg{\con{s}}( a ) \leq 0,\;
          \con{s}(\farg{\con{s}}( a )) \teq 0,\;
\\[-\jot]
  & \con{s}(\farg{\con{s}}( a )) \teq \farg{\con{s}}( a ) + \con{s}( \farg{\con{s}}( a )-1 )
    \wedge \exists b : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1 \bigr) \bigr)
\end{aligned}
\\[-\jot]
 & {\wedge}\; \con{s}( \con{c} ) > 100 \wedge \exists a : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( a ) \teq \con{c}
\end{aligned}
\end{equation}
%
The universal quantification in the resulting formula is over an uninterpreted
type $\fargtype{\con{s}}$, thus enabling the finite model finding
techniques by Reynolds et al.\ \cite{ReyEtAl-1-RR-13,reynolds-et-al-2013},
implemented in \cvc, to search for a finite interpretation for $\fargtype{\con{s}}$. 
Since all occurrences of the quantified variable~$a$ in this formula are 
beneath applications of the uninterpreted function $\farg{\con{s}}$, 
the formula is in the essentially uninterpreted fragment,
for which Ge and de Moura \cite{GeDeM-CAV-09} provide 
a complete instantiation procedure, implemented in \ziii. 
Accordingly,
%%% Added "apparently". Or are we sure?  
% Yes, I'm planning to post the 2 versions of the benchmark on the evaluation page (when we add it) --ajr
% Yes, but how can you be sure that Z3 will run forever? Your benchmarks surely have a timeout! --jb
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), 
whereas they produce a model for formula~(\ref{eq:ex-after}) 
within 100 milliseconds.\xend
\end{example}

>>>>

Our setting is a monomorphic (or many-sorted) first-order logic
like the one defined by SMT-LIB \cite{smtlib25}.
A \emph{signature} $\Sigma$ consists of
a set $\stypes{\Sigma}$ of %ct type symbols
first-order types
%
%\ct{What the rationale for using the letter $y$ in $\stypes{\Sigma}$?}
%\jb{y as in t\textbf{y}pe. changed to `ty'}
(or sorts) and a set $\sfuns{\Sigma}$ of function symbols over these types.
We assume that signatures always contain a Boolean type $\Bool$ and constants
$\ltrue, \lfalse : \Bool$ for truth and falsity,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \stypes{\Sigma}$,
standard Boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and an if--then--else function symbol
$\lite : \Bool \times \tau \times \tau \rightarrow \tau$
for each $\tau \in \stypes{\Sigma}$.
%
For each $\tau \in \stypes{\Sigma}$,
we fix an infinite set $\svars{\Sigma}_\tau$ of \emph{variables of type $\tau$} and
define $\svars{\Sigma}$ as $\bigcup_{\tau \in \stypes{\Sigma}} \svars{\Sigma}_\tau$.
$\Sigma$-terms are built as usual over functions symbols in $\Sigma$ and variables in $\svars{\Sigma}$.
%
Formulas are terms of type $\Bool$.
We write $t\typ{\tau}$ to denote %ct a metavariable ranging over 
terms of type~$\tau$ and $\terms( t )$ to denote the set of subterms in $t$.
Given a term $u$, we write $u[\vec t/\vec x]$ to denote the result of replacing
all occurrences of $\vec x$ with $\vec t$ in $u$.
\begin{longv}When applied to terms, the symbol $=$ denotes syntactic equality.\end{longv}

@@

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well-founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)$,
where $T$ is integer arithmetic extended with the uninterpreted symbol 
$\con p:\typeint \rightarrow \typeint$, can be shown well-founded under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. \begin{longv}Logically, such dependencies can be captured by
congruence rules. \end{longv}Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

@@

%\newcommand\badassex{
% \{\forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x ),\;\,
% \forallf{\con{g}} x : \typeint.\;\, \lite( \con{f}( x ) \teq 5,\allowbreak\;
%   \con{g}( x ) \teq \con{f}( x ),\allowbreak\;
%   \con{g}( x ) \teq \con{g}( x ) + 1)\}}
\newcommand\badassex{
 \{\forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x ),\allowbreak\;\,
 \forallf{\con{g}} x : \typeint.\;\, \con{g}( x ) \teq \con{g}( x ) + \nobreak\con{f}( x )\}}

An example of an inadmissible set is
$\{ \forallf{\con{f}} x : \typeint.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$,
where $T$ is integer arithmetic extended to a set of uninterpreted symbols 
$\{\con f, \con g :\typeint \rightarrow \typeint,\, \ldots \}$.
The reason is that the formula $\ltrue$ is (trivially) closed under function expansion with respect to this set,
and there is no model of $T$
satisfying $\con{f}$'s definition. A more subtle example is
\begin{shortv}$\badassex$. \end{shortv}%
\begin{longv}\[\badassex\]\end{longv}%
While this set has a model where $\con f$ and $\con g$ are interpreted as the
constant function $0$, it is not admissible since %the formula
$\con{f}( 0 ) \teq 1$ is
closed under function expansion \begin{longv}with respect to this set\end{longv}
and yet there exists
no interpretation satisfying both $\con{f}( 0 ) \teq 1$ and $\con{g}$'s
definition.

@@

\begin{figure}[t]
\normalsize
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv_0( t\typ{\tau},\, p )$ $=$
%\\[-.8\baselineskip]
 \begin{itemize}
   \itemx[] $\mathsf{if}$ $\tau = \Bool$ $\mathsf{and}$ $t = \boolop(t_1,\ldots,t_n)$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( t'_i,\, \chi_i ) = \conv_0( t_i,\, \pol( \boolop,\, i,\, p ) )$ $\mathsf{for}$ $i = 1, \ldots, n$ $\mathsf{in}$%\kern6.35em$\mid$
      \itemx[] $\mathsf{let}$ $\chi = \chi_1 \wedge \cdots \wedge \chi_n$ $\mathsf{in}$
      \itemx[] $\mathsf{if}$ $p = \ppos$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) ) \wedge \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ $\mathsf{if}$ $p = \pneg$ $\mathsf{then}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n) \vee \neg \chi,\, \ltrue$}
      \itemx[] $\mathsf{else}$ \bigtuple{$\boolop(t'_1, \ldots, t'_n),\, \chi$}
%\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forallf{\con{f}} \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall a : \fargtype{\con{f}}.\; u' [ \vecfarg{\con{f}}( a ) / \vec x ],\, \ltrue$}
%\\[-.8\baselineskip]
    \end{itemize}
  \itemx[] $\mathsf{else}$ $\mathsf{if}$ $t = \forall \vec x.\;\, u$ $\mathsf{then}$
    \begin{itemize}
      \itemx[] $\mathsf{let}$ $( u',\, \chi ) = \conv_0( u,\, p )$ $\mathsf{in}$ \bigtuple{$\forall \vec x.\; u',\, \forall \vec x.\; \chi$}
%\\[-.8\baselineskip]
    \end{itemize}
   \itemx[] $\mathsf{else}$
   \begin{itemize}
     \itemx[] \return \bigtuple{$t,\; \bigwedge\, \{ \exists a : \fargtype{\con{f}}.\; \vecfarg{\con{f}}( a ) \teq \vec s \mid \con{f}( \vec s ) \in \terms( t ),\, \con{f} \in \sfundefs{\Sigma} \}$}
%\\[-.8\baselineskip]
   \end{itemize}
 \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\itemx[\ ]
$\conv( \varphi )$ $=$ $\mathsf{let}$ $( \varphi',\, \chi ) = \conv_0( \varphi,\, \ppos )$ $\mathsf{in}$ $\varphi'$
%\\[-1.25\baselineskip] %% TYPESETTING
\end{enumerate}
\caption{\,Definition of translation $\conv$}
\label{fig:A}
\end{figure}

For the rest of the section, let $\varphi$ be a $\Sigma$-formula
in definitional form with respect to $\sfundefs{\Sigma}$
whose definitions are admissible.
%We may translate $\varphi$ into a equisatisfiable formula $\phi'$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
We present a method that constructs an extended signature
$\extendsig{ \Sigma }$ and an $\extendsig{ \Sigma }$-formula $\varphi'$ such that
$\varphi'$ is $T$-satisfiable if and only if $\varphi$ is $T$-satisfiable---i.e.,
$\varphi$ and $\varphi'$ are \emph{equisatisfiable} (\emph{in $T$}).
The idea behind this translation
is to use an uninterpreted type to abstract the set of
\emph{relevant} tuples for each defined function $\con{f}$ and restrict the
quantification of $\con{f}$'s definition to a variable of this
type. 
Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the
ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the
satisfiability of $\varphi$.
More precisely,
%suppose our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
for each $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau \in \sfundefs{\Sigma}$,
the extended signature $\extendsig{\Sigma}$ contains
%\begin{itemize}
%\item
an uninterpreted \emph{abstract type} $\fargtype{\con{f}}$ and
%\item
$n$ uninterpreted \emph{concretization functions} $\fargx{\con{f}}{1} : \fargtype{\con{f}} \rightarrow \tau_1$, \ldots, $\fargx{\con{f}}{n} : \fargtype{\con{f}} \rightarrow \tau_n$.
%\end{itemize}
%The interpretation of uninterpreted type $\fargtype{\con{f}}$ will denote the elements (tuples) on which the function $\con{f}$ is applied.
%The role of the uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ will be discussed more in the following.

The translation $\conv$ defined in Figure~\ref{fig:A} translates the $\Sigma$-formula
$\varphi$ into the $\extendsig{\Sigma}$-formula $\varphi'$. It relies
on the auxiliary function $\conv_0$, which takes two arguments:\ the term $t$
to translate and a \relax{polarity}~$p$ for $t$, which is either $\ppos$, $\pneg$, or
$\pnone$. 
\vthinspace$\conv_0$ returns a pair $( t'\negvthinspace,\, \chi )$, where $t'$ is a term of
the same type as $t$ and $\chi$ is an $\extendsig{\Sigma}$-formula.
%an additional constraint.
%This constraint is needed if the polarity is $\pnone$; otherwise, it is encoded
%directly in $t'$ and $\chi$ is simply $\ltrue$.
%In the auxiliary function $\conv_0$, the function $\pol(
%\boolop, i, p )$ returns the polarity of the $i$\vvthinspace{th} argument of an
%application of~$\boolop$ having polarity $p$ if $\boolop$ is an interpreted
%predicate or $\pnone$ otherwise.

@@

The translation alters the formula $\varphi$ in two ways. First, it restricts the
quantification on function definitions for $\con{f}$ to the corresponding
uninterpreted type $\fargtype{\con{f}}$, inserting applications of the concretization functions $\fargx{\con{f}}{i}$ as needed. 
Second, it augments $\varphi$ with additional constraints of the form
$\exists a :\nobreak \fargtype{\con{f}}.\;\, \vecfarg{\con{f}}( a ) \teq \vec s$,
where $\vecfarg{\con{f}}( a ) \teq \vec s$ abbreviates the formula
$\bigwedge_{i=1}^n\, \fargx{\con{f}}{i}(a) \teq s_i$
with $\vec s = (s_1,\ldots,s_n)$.
These existential %%% for typesetting
constraints
ensure that the restricted definition for $\con{f}$ covers all relevant tuples
of terms, namely those occurring in applications of $\con{f}$ 
that are relevant to the satisfiability of $\varphi$.

If $t$ is an application of a predicate symbol $\boolop$, including the
operators ${\oldneg}$, ${\wedge}$, ${\vee}$, ${\teq}$, and ${\lite}$,
$\conv_0$ calls itself recursively on the arguments $t_i$ and polarity $\pol(
\boolop, i, p )$, with $\pol$ defined as
\[\pol( \boolop, i, p ) =
\begin{cases}
p & \text{if either $\boolop \in \{{\wedge}, {\vee}\}$ or $\boolop = \lite$ and $i \in \{2, 3\}$} \\[-\jot]
\pnegate p & \text{if $\boolop = {\neg}$} \\[-\jot]
\pnone & \text{otherwise}
\end{cases}\]
where $\pnegate p$ is $\pneg$ if $p$ is $\ppos$, $\ppos$
if $p$ is $\pneg$, and $\pnone$ if $p$ is $\pnone$.
The term $t$ is then reconstructed as $\boolop(t_1',\ldots,t_n')$ 
where each $t_i'$ is the result of the recursive call with argument $t_i$. 
If the polarity $p$ associated with $t$ is $\ppos$, 
$\conv_0$ conjunctively adds to $\boolop(t_1',\ldots,t_n')$ the constraint $\chi$ derived from
the subterms. Dually, if $p$ is $\pneg$, it adds a
disjunction with the negated constraint, to achieve the same overall effect.
%ct  as in the positive case. 
It $p$ is $\pnone$, the constraint $\chi$ is
returned to the caller.
%
\begin{longv}\par\end{longv}
%
If $t$ is a function definition, % of a function in $\sfundefs{\Sigma}$,
$\conv_0$ constructs a quantified formula over a single variable
$a$ of type $\fargtype{\con{f}}$ and replaces all occurrences of $\vec x$ in
the body of that formula with $\vecfarg{\con{f}}( a )$. 
(Since function definitions are top-level conjuncts, %by case analysis on the return values of $\conv$,
$\chi$ must be $\ltrue$ and can be ignored.)
%
If $t$ is an unannotated quantified formula, $\conv_0$ calls itself
on the body with the same polarity. \begin{longv}A quantifier is prefixed to the
quantified formula and constraint returned by the recursive call.\end{longv}
Otherwise, $t$ is either an application of an uninterpreted predicate symbol or a term
of a type other than $\Bool$. Then, the returned constraint is a conjunction of
formulas of the form $\exists a : {\fargtype{\con{f}}}.\;\, \vecfarg{\con{f}}(
a ) \teq \vec s$ for each subterm $\con{f}( \vec s )$ of~$t$ such that $\con{f}
\in \sfundefs{\Sigma}$. Such constraints, when asserted positively, ensure that
some element in the abstract domain $\fargtype{\con f}$ is the preimage of
the argument tuple $\vec s$. % under $\vecfarg{\con{f}}$.

@@

\begin{example}
\label{ex:translation}
Let $T$ be linear arithmetic with the uninterpreted symbols  $\{ \const{c} : { \typeint },\> \con{s} : { \typeint \rightarrow \typeint } \}$.
Let $\varphi$ be the $\Sigma$-formula
\begin{equation} \label{eq:ex-before}
\forallf{\con{s}} x : {\typeint}.\;\, \lite\bigl( x \leq 0,\; \con{s}(x) \teq 0,\;
  \con{s}( x ) \teq x + \con{s}( x - 1 ) \bigr) \wedge \con{s}( \con{c} ) > 100
\end{equation}
%
The definition of $\const{s}$ specifies that it returns the sum of all
positive integers up to $x$. The formula $\varphi$ is in definitional form with
respect to $\sfundefs{\Sigma}$ 
%\jb{correct? It said $\Sigma_u$, which was definitely wrong} 
and states that the sum of all
positive numbers up to some constant $\const{c}$ is greater than $100$. It is
satisfiable with a model that interprets $\const{c}$ as $14$ or more.
Due to the universal quantifier,
current SMT techniques
are unable to find
a model for $\varphi$. The signature $\extendsig{\Sigma}$ extends $\Sigma$ with the type
$\fargtype{\con{s}}$ and the uninterpreted function symbol $\farg{\con{s}} : \fargtype{\con{s}}
\rightarrow \typeint$. The result of $\conv( \varphi )$, after simplification, 
is the $\extendsig{\Sigma}$-formula
%
\begin{equation} \label{eq:ex-after}
\!\begin{aligned}[c]
  & \phantom{{\wedge}\; }\bigl(
      \forall a : \fargtype{\con{s}}.\; \lite\bigl(
        \!\begin{aligned}[t]
         &  \farg{\con{s}}( a ) \leq 0,\;
          \con{s}(\farg{\con{s}}( a )) \teq 0,\;
\\[-\jot]
  & \con{s}(\farg{\con{s}}( a )) \teq \farg{\con{s}}( a ) + \con{s}( \farg{\con{s}}( a )-1 )
    \wedge \exists b : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( b ) \teq \farg{\con{s}}( a )-1 \bigr) \bigr)
\end{aligned}
\\[-\jot]
 & {\wedge}\; \con{s}( \con{c} ) > 100 \wedge \exists a : {\fargtype{\con{s}}}.\;\, \farg{\con{s}}( a ) \teq \con{c}
\end{aligned}
\end{equation}
%
The universal quantifier in formula~(\ref{eq:ex-after}) ranges over an uninterpreted
type $\fargtype{\con{s}}$, making it amenable to the finite model finding
techniques by Reynolds et al.\ \cite{ReyEtAl-1-RR-13,reynolds-et-al-2013},
implemented in \cvc, which search for a finite interpretation for $\fargtype{\con{s}}$. 
Furthermore, since all occurrences of the quantified variable~$a$ are 
beneath applications of the uninterpreted function $\farg{\con{s}}$, 
the formula is in the essentially uninterpreted fragment,
for which Ge and de Moura \cite{GeDeM-CAV-09} provide 
a complete instantiation procedure, implemented in \ziii. 
As expected,
%%% Are we sure?  
% Yes, I'm planning to post the 2 versions of the benchmark on the evaluation page (when we add it) --ajr
% Yes, but how can you be sure that Z3 will run forever? Your benchmarks surely have a timeout! --jb
\cvc and \ziii run indefinitely on formula~(\ref{eq:ex-before}), 
whereas they produce a model for~(\ref{eq:ex-after}) 
within 100 milliseconds.\xend
\end{example}
